Variance reduction for Markov chains via martingale representations

### 2020-05-09 aap/Revision/Response_1.tex

We thank the referee for carefully reading our manuscript and for pertinent
comments which helped us to improve the paper. In the new version of our
paper, we have taken all comments and suggestion into account.  Below are some additional responses to the
comments by Referee 1

\section*{Main remarks}

\begin{itemize}

\item \textit{... The authors make an attempt to quantify the variance reduction obtained by the proposed approach, as well as the computational complexity to achieve this variance reduction. On these topics several results are presented as Theorems and Propositions  ... I would recommend the authors to phrase this results with clear conditions in such a way that it becomes directly accessible to the reader ... where the bound is phrased in an easily readible way involving constants that are defined in the statement of the theorem and which should not depend on infinite series and conditional expectations.}
\par
Let us first stress that  current literature on MCMC  mainly focuses on  convergence of MCMC algorithms in Total Variation and Wasserstein distances thus  ignoring a sample error of these algorithms. Only few references studied MSE error of the Langevin algorithm (see e.g.  Durmus and  Moulines (2017)). In this work we propose a novel variance reduction approach which is able to reduce the statistical error of MCMC algorithms. The main advantage of our method is provable increase of efficiency for Langevin-based MCMC algorithms as measured by  the cost-to-variance criterion. Our approach is based on a new discrete time martingale representation for Markov chains which is of its own interest. Therefore we first present this basic result and then discuss in Section 4 variance reduction method based on the derived martingale representation. In  Section~4 we  explain in accessible way why our approach leads to a smaller cost-to-variance characteristic. These results are based on deep analysis of the variance reduced estimate postponed to Section 5. In the revised version we significantly weaken  assumptions  and prove our main result (Theorem~10 and Corollary~11) for ULA under smoothness  and convexity of the underlying potential outside a ball (assumption  (H2)).   Note that all assumptions needed are now collected in Theorem~10. For convenience of the reader we also sketch the main steps of the proof right after 
Theorem~10. We decided to present our main theoretical result in the form of bounds for the functions $A_{q,k}$ since these functions play a crucial role in the variance bound (13). Given these estimates it is easy to see how to get bounds for the variance. 
\item \textit{... It is good to see that the technique has beneficial effects on the variance, but it does not become clear what was the computational effort required to achieve this benefit. The reader is left wondering whether or not the same variance reduction could have been achieved simply by drawing more MCMC samples ...}
\item \textit{... here the setting \(\gamma_k = c k^{-\alpha}, \alpha\in  (0, 1)\) is considered, and presented as  ``the most interesting case''. For many practitioners, $\alpha=0$ would be rather natural. What can be said in this important case? This is also applies to Section 6 ...}
\par
In the revised version we decided to switch to a constant time step and present all our bounds for this case. On the one hand, this allows us to formulate our main results in a simpler form depending only on a single parameter $\gamma.$ On the other hand, this is indeed the most relevant choice for practitioners.  
\item \textit{ ... here the functions g are assumed uniformly bounded, whereas earlier (end of Section 5) elements in $\Psi$ were proposed to be polynomials. In particular the bound B would be equal to infinity. How can this be reconciled? ...}
\par
In the revised version we skip this result because of size restrictions. 
\end{itemize}

### 2020-05-09 aap/Revision/Response_2.tex

We thank the referee for carefully reading our manuscript and for pertinent
comments which helped us to improve the paper. In the new version of our
paper, we have taken all comments and suggestion into account.  Below are some additional responses to the
comments by Referee 2

\section*{Main remarks}

\begin{itemize}

\item \textit{... Restrictive assumptions for the theoretical analysis. The main theorems of the paper, which seem to be Theorem 4 and the results of section 6.1, have the restrictive assumption that the drift vector field  needs to satisfy the strong convexity assumption ... In typically interesting cases, such as Bayesian inverse problems and machine learning, this assumption is not satisfied. Unfortunately, even in the numerical examples provided in Section 7, this assumption is not fulfilled ? it already doesn't hold for a two-component Gaussian mixture in one dimension. This is a serious restriction ... }
\par
In the revised version we significantly weaken  these assumptions and prove our main result (Theorem~10 and Corollary~11) for ULA under smoothness  and convexity of the underlying potential outside a ball (assumption  (H2)) hence avoiding the strong convexity assumption.  Note that all assumptions needed are now collected in Theorem~10 and Corollary~11. These  assumptions can now be validated for a large class of examples including the case of Gaussian mixtures considered in Section~6.2.
\item \textit{... in complicated models such as PDE-related Bayesian inverse problems or machine learning, the computation of \(\nabla \log(\pi)\) is expensive ? but these examples are typically ones where \(\log(\pi)\) does not satisfy the needed concavity assumptions ...}
\par
As mentioned above we have replaced the strong convexity assumption by a much weaker one (H2) (convexity outside a ball). Hence our method may gain additional efficiency as compared to other variance reduction approaches due to the fact that \(\nabla \log(\pi)\) doesn't need to be computed for the construction of control variates.
\end{itemize}

### 2020-05-09 aap/Revision/vr-mcmc-revision-final.tex
update

X_{k+1} = (1-2\gamma)X_k + \sqrt{\gamma}\xi_{k+1},\quad \xi_{k+1} \sim \mathcal{N}(0,1).

@@

= \frac{1}{n^2}\PVar\left[\sum\limits_{i=1}^{n}\sum\limits_{j=1}^{i}\sqrt{\gamma}(1-2\gamma)^{i-j}\xi_{j}\right]
\geq  \frac{1}{n\gamma} - \frac{2(1-2\gamma)}{n^2\gamma^2}.

@@

\max_{j\geq N}\sum_{p=j+1}^{\infty}\gamma\,\left\Vert\prod_{k=j+1}^{p} (I-\gamma\mathsf{D}^{2}U(X^x_{k-1}))\right \Vert_{r}\leq \Xi,\quad r=2,4,

@@

for some $\Xi=\Xi(x)>0$ not depending on $\gamma,$ where for any random matrix $A$ we define $\|A\|_r=(\mathsf{E}|A|^r)^{1/r}.$ 

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{../pictures/logreg_1st_order.png}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Logistic Regression. The compared estimators are the ordinary empirical average (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) control variates obtained with empirical spectral variance minimisation (ESVM).
 \label{fig:2}}
\end{center}
\end{figure}

In the case of  Gaussian prior $\pi_0(\theta) \sim \mathcal{N}(0,\sigma^2 I_d)$, the unnormalized posterior density takes the form:
\begin{eqnarray*}
\pi(\theta | \mathsf{Y}, \mathsf{X}) \propto \exp \left\{ \mathsf{Y}^T \mathsf{X} \theta - \sum\limits_{i=1}^m \ln{\left(1 + \rme^{\mathsf{X}_i \theta}\right)} - \frac{1}{2\sigma^2} \left\|\theta\right\|_2^2 \right\},
\end{eqnarray*}
Thus we obtain
\begin{eqnarray*}
U(\theta) &=& -\mathsf{Y}^T \mathsf{X} \theta + \sum_{i=1}^m \ln{\left(1 + \rme^{\theta^T \mathsf{X}_i}\right)} + \frac{1}{2\sigma^2} \left\| \theta\right\|_2^2,
\\
\nabla U(\theta) &=& -\mathsf{Y}^T \mathsf{X} + \sum_{i=1}^m \frac{\mathsf{X}_i}{1 + \rme^{-\mathsf{\theta}^T \mathsf{X}_i}} + \frac{1}{\sigma^2} \theta.
\end{eqnarray*}
To demonstrate the performance of the proposed control variates approach in the above Bayesian logistic regression model, we take a simple dataset from \cite{mira2013zero}, which contains the measurements of four variables on $m=200$ Swiss banknotes. Prior distribution of the regression parameter $\mathbf{\theta} = \left( \theta_1, \theta_2, \theta_3, \theta_4 \right)$ assumed to be normal with the covariance matrix $\sigma^2 I_4$, where $\sigma^2 = 100$. To construct trajectories of length $n = 1 \times 10^4, $ we take step size $\gamma = 0.1$ for the ULA scheme with $N = 10^3$ burn-in steps. As in the previous experiment we use the first order polynomials approximations to analytically compute the coefficients $\hat{a}_{p-l,\mathbf{k}},$ based on $T = 10$ training trajectories. We put $n_{trunc} = 50$ and $K = 1$. The target function is taken to be $f(\theta) = \sum_{i=1}^{i=d}\theta_i$.  In order to test our variance reduction algorithm, we generate \(100\) independent test trajectories of length $n = 10^3$. In Figure ~\ref{fig:2} we compare our approach to the variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use first order polynomials for the same reasons as in the Gaussian mixtures example.

@@

\begin{proof} Note that the condition {\bf (H1)} implies that for any $x,y \in \rset^d$,
\[
|x-y-\gamma(\nabla U(x) - \nabla U(y))|^2 \leq (1 + 2L\gamma + \gamma^2)|x-y|^2
\] 
Hence,   \cite[Corollary 5]{DurmusBortoli} implies that for all $x,y$ such that $|x|,|y| \leq K$,
\[
\|\delta_xQ^{\lceil 1/\gamma \rceil}_{\gamma} - \delta_yQ^{ \lceil 1/\gamma \rceil}_{\gamma}\|_{\text{TV}} \leq 2(1-\varepsilon)
\]
where $\lceil \cdot \rceil$ denotes an upper integer part and $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$. Note also that the drift condition \eqref{eq:lyapunov} implies that
\[
Q^{[1/\gamma]}_{\gamma}V(x) \leq \lambda V(x) + \frac{\gamma b(1-\lambda^{\gamma(\lceil 1/\gamma\rceil + 1)})}{1-\lambda^{\gamma}} \leq \lambda V(x) + \frac{b}{\lambda \log{1 /\lambda}} 
\]
where the last inequality is due to \cite[Lemma 1]{durmus:moulines:2017}. Hence, the kernel $Q^{\lceil 1/\gamma \rceil}_{\gamma}$ satisfies $(1,\varepsilon)$-Doeblin condition on the set $\{x: V(x) < 1+K^2\}$ and drift condition \eqref{eq:lyapunov} with appropriate constants. Now the statement of the lemma follows from \cite[Theorem~19.4.1]{moulines2018}.
\end{proof}

>>>>

X_{k+1} = \left(1-\frac{\gamma}{2}\right)X_k + \sqrt{\gamma}\xi_{k+1},\quad \xi_{k+1} \sim \mathcal{N}(0,1).

@@

= \frac{1}{n^2}\PVar\left[\sum\limits_{i=1}^{n}\sum\limits_{j=1}^{i}\sqrt{\gamma}\left(1-\frac{\gamma}{2}\right)^{i-j}\xi_{j}\right]
\geq  \frac{4}{n\gamma} - \frac{16(1-\frac{\gamma}{2})}{n^2\gamma^2}.

@@

\max_{j\geq N}\sum_{p=j+1}^{\infty}\gamma\,\left\Vert\prod_{k=j+1}^{p} (I-\gamma\mathsf{D}^{2}U(X^x_{k-1}))\right \Vert_{r}\leq \Xi(x),\quad r=2,4,

@@

for some $\Xi(x)>0$ not depending on $\gamma,$ where for any random matrix $A$ we define $\|A\|_r=(\mathsf{E}|A|^r)^{1/r}.$ 

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{../pictures/logreg_1st_order.png}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Logistic Regression. The compared estimators are the ordinary empirical average (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) control variates obtained with empirical spectral variance minimisation (ESVM).
 \label{fig:2}}
\end{center}
\end{figure}
In the case of  Gaussian prior $\pi_0(\theta) \sim \mathcal{N}(0,\sigma^2 I_d)$, the unnormalized posterior density takes the form:
\begin{eqnarray*}
\pi(\theta | \mathsf{Y}, \mathsf{X}) \propto \exp \left\{ \mathsf{Y}^T \mathsf{X} \theta - \sum\limits_{i=1}^m \ln{\left(1 + \rme^{\mathsf{X}_i \theta}\right)} - \frac{1}{2\sigma^2} \left\|\theta\right\|_2^2 \right\},
\end{eqnarray*}
Thus we obtain
\begin{eqnarray*}
U(\theta) &=& -\mathsf{Y}^T \mathsf{X} \theta + \sum_{i=1}^m \ln{\left(1 + \rme^{\theta^T \mathsf{X}_i}\right)} + \frac{1}{2\sigma^2} \left\| \theta\right\|_2^2,
\\
\nabla U(\theta) &=& -\mathsf{Y}^T \mathsf{X} + \sum_{i=1}^m \frac{\mathsf{X}_i}{1 + \rme^{-\mathsf{\theta}^T \mathsf{X}_i}} + \frac{1}{\sigma^2} \theta.
\end{eqnarray*}
To demonstrate the performance of the proposed control variates approach in the above Bayesian logistic regression model, we take a simple dataset from \cite{mira2013zero}, which contains the measurements of four variables on $m=200$ Swiss banknotes. Prior distribution of the regression parameter $\mathbf{\theta} = \left( \theta_1, \theta_2, \theta_3, \theta_4 \right)$ assumed to be normal with the covariance matrix $\sigma^2 I_4$, where $\sigma^2 = 100$. To construct trajectories of length $n = 1 \times 10^4, $ we take step size $\gamma = 0.1$ for the ULA scheme with $N = 10^3$ burn-in steps. As in the previous experiment we use the first order polynomials approximations to analytically compute the coefficients $\hat{a}_{p-l,\mathbf{k}},$ based on $T = 10$ training trajectories. We put $n_{trunc} = 50$ and $K = 1$. The target function is taken to be $f(\theta) = \sum_{i=1}^{d}\theta_i$.  In order to test our variance reduction algorithm, we generate \(100\) independent test trajectories of length $n = 10^3$. In Figure ~\ref{fig:2} we compare our approach to the variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use first order polynomials for the same reasons as in the Gaussian mixtures example.

@@

\begin{proof} Note that the condition {\bf (H1)} implies that for any $x,y \in \rset^d$,
\[
|x-y-\gamma(\nabla U(x) - \nabla U(y))|^2 \leq (1 + 2L\gamma + \gamma^2)|x-y|^2
\] 
Hence,   \cite[Corollary 5]{DurmusBortoli} implies that for all $x,y$ such that $|x|,|y| \leq K$,
\[
\|\delta_xQ^{\lceil 1/\gamma \rceil}_{\gamma} - \delta_yQ^{ \lceil 1/\gamma \rceil}_{\gamma}\|_{\text{TV}} \leq 2(1-\varepsilon)
\]
where $\lceil \cdot \rceil$ denotes an upper integer part and $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$. Note also that the drift condition \eqref{eq:lyapunov} implies that
\[
Q^{\lceil 1/\gamma \rceil}_{\gamma}V(x) \leq \lambda V(x) + \frac{\gamma b(1-\lambda^{\gamma(\lceil 1/\gamma\rceil + 1)})}{1-\lambda^{\gamma}} \leq \lambda V(x) + \frac{b}{\lambda \log{1 /\lambda}} 
\]
where the last inequality is due to \cite[Lemma 1]{durmus:moulines:2017}. Hence, the kernel $Q^{\lceil 1/\gamma \rceil}_{\gamma}$ satisfies $(1,\varepsilon)$-Doeblin condition on the set $\{x: V(x) < 1+K^2\}$ and drift condition \eqref{eq:lyapunov} with appropriate constants. Now the statement of the lemma follows from \cite[Theorem~19.4.1]{moulines2018}.
\end{proof}
